# Assignment 3 data analysis

```{r}
date()
```

```{r}
library(tidyverse)
library(ggplot2)
library(finalfit)
library(broom)
library(boot)
theme_set(theme_bw())
```

**2. reading the data**

```{r}
student_data <- read_csv("https://raw.githubusercontent.com/KimmoVehkalahti/Helsinki-Open-Data-Science/master/datasets/alc.csv")

glimpse(student_data)
```

The dataset includes data from two Portuguese schools. It includes data for 370 students (rows) and 35 different attributes (columns). It has data for two student demographics such as sex, age and home address type (urban or rural). For student performance it includes data from Math and Portuguese classes from three grades.

##Exploring student alcohol consumption

**3. setting the study hypothesis**

I decided to study how sex, address, paid, and absences may affect alcohol use. My hypothesis is that males use more alcohol than females. 
*address* variable tells if the student's home address is rural (R) or urban (U), and I think that there may be differences between urban and rural locations. 
*paid* variable tell if student paid for extra classes in either of the subjects (math or portugal). My hypothesis is that those who paid for extra classes drink less alcohol. 
*absences* is the number of student's school absences. My hypothesis is that those who drink more alcohol have more absences.

```{r}
#creating dataset with only the selected variables
model_data <- student_data %>% 
  select(sex, address, paid, absences, alc_use, high_use)
  
glimpse(model_data)
```
**4. Exploration of the data**

*Graphical exploration*

```{r}
model_data %>% 
  ggplot(aes(x = high_use))+
  geom_bar(colour = "blue", fill="lightblue")+
  facet_wrap(~sex)+
  labs(x="High alcohol use", title = "Alcohol use by sex")
```

Based on the visual inspection, it seems that high alcohol use is more common among men.

```{r}
model_data %>% 
  ggplot(aes(x = high_use))+
  geom_bar(colour = "forestgreen", fill="lightgreen")+
  facet_wrap(~address)+
  labs(x="High alcohol use", title="Rural (R) and urban (U)")
```

It seems that high alcohol use is proportionally more common in rural areas than in urban areas.

```{r}
model_data %>% 
  ggplot(aes(x = high_use))+
  geom_bar(colour = "orange", fill="yellow")+
  facet_wrap(~paid)+
  labs(x="High alcohol use", title = "Paid for extra classes")
```
No clear difference is seen between those who paid for extra classes and those did not. There seems to be weak tendency towards high alcohol use in those who paid for extra classes.


```{r}
model_data %>% 
  ggplot(aes(y = high_use, x=absences))+
  geom_boxplot(colour = "red", fill="coral")+
  labs(y="High alcohol use", x = "Absences", title = "High alcohol use and school absences")
```
It seems that high alcohol use correlates with school absences.

*Numerical exploration*

```{r}
model_data %>% 
  summary_factorlist("high_use", c("sex", "address", "paid", "absences"),
                     cont = "mean", p = T, add_dependent_label = TRUE)
```
As was seen in the graphical exploration, males have statistically higher portion of high use of alcohol. It seems that the hypothesis was correct on this part. 

There is no statistical significance in high alcohol between urban and rural locations. However, it seems that there is tendency towards higher alcohol use in rural areas. Hypothesis may or may not be correct.

There is no difference in alcohol between those who paid or did no pay for extra classes.

Absences correlate with high use of alcohol as per hypothesis. The difference is statistically significant.

**5. logistic regression**

```{r}

fit1 <- glm(high_use~sex+address+paid+absences, family = binomial, data =model_data)

summary(fit1)

```
```{r}
fit1 %>% 
  tidy(conf.int=T, exp=T) #calculate confidence intervals and transfrom back from the natural logarithm with exp() 
```
In the table above, estimate column tells the odd ratios, while conf.low is the lower limit and conf.high the upper limit of 95% confidence interval.

In the logistic regression model sex and absences had statistically significant association.

Hypothesis regarding sex was that high use of alcohol is more common among males. The logistic regression model supports this hypothesis and the odd ratio for high alcohol use is 2.95.

Another hypothesis was that there is difference between rural and urban areas. The difference is not statistically significant, but the p-value is quite small (0.08), and there seems to be at least trend that alcohol use is less common in urban areas (odd ratio: 0.62, 95% confidence interval: 0.36-1.07).

Hypothesis for association between paid classes and alcohol use does not seem to be correct. There was no difference according to this analysis.

As per hypothesis, absences correlated with high alcohol use. P-value was small (2.9x10^-9^). For the absences the odd ratio was 1.1 and 95% confidence interval 1.06-1.16).

**6. **

```{r}
fit2 <- glm(high_use~sex+absences, family = binomial, data =model_data) #new model with only statistically significant variables

summary(fit2)
```

```{r}
probabilities <- predict.glm(fit2, type = "response") #predict probabilites based on the glm-model

model_data <- model_data %>% 
    mutate(probability = probabilities,
           prediction = probability>0.5) #create prediction variable, which is TRUE when probability > 0.5

prediction_table <- table(high_use = model_data$high_use, prediction = model_data$prediction)
prediction_table

```

```{r}
training_error <- (prediction_table[1,2] + prediction_table[2,1])/370 #proportion of correctly classified values
training_error

v1 = c("word", "ok")

model_data %>% 
  mutate(correct = high_use==prediction) %>% 
  ggplot(aes(x = high_use, y=prediction, colour=correct))+
  geom_point()+
  geom_jitter()+
  labs(colour = "Correct prediction", title = "Actual values vs. prediction (training error = 0.25)")
```
Model has training error of 25%. It has tendency towards underprediction of alcohol use. It predicts around 3% of false positives (8/259), but large amount of false negatives, around 77% (86/111). I would not say that this model is actually useful.

```{r}
easy_prediction <- model_data %>% 
  mutate(high_absences = absences>6) #lets compare how well over 6 absences predict high alcohol use

pred_table2 <- table(high_use = easy_prediction$high_use, prediction = easy_prediction$high_absences)

pred_table2
```

If we take more simple approach and predict that those who have over 6 absences have high alcohol use we get more true positives and less true negative.

```{r}
(pred_table2[1,2] + pred_table2[2,1])/370 
```
Training error of the simpler model is 31% which is a little bit worse than the models training error (25%).

**7. cross-validation**
```{r}
loss_func <- function(class, prob) {
  n_wrong <- abs(class - prob) > 0.5
  mean(n_wrong)
}


# K-fold cross-validation
cv <- cv.glm(data = model_data, cost = loss_func, glmfit = fit2, K = nrow(model_data))

# average number of wrong predictions in the cross validation
cv$delta[1]


```

